---
title: "age_predictions"
author: "Anderson Butler"
date: "2023-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(readr)
# load data
df_normalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221111_dnamstability_normalizedBetas_sesame_abutler.Rds")
df_unnormalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221111_dnamstability_unnormalizedBetas_sesame_abutler.Rds")
df_normalized_with_combat = read_csv("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_fluorescence_values/sesameAndCombat_normalized_betas_for_kras_20230317.csv")


df_masking_test = read_csv("df_masking_CV_point75.csv")
df_masking_test = data.frame(df_masking_test)
rownames(df_masking_test) = df_masking_test$cpgs
df_masking_test$cpgs = NULL
df_masking_test$...1=NULL

df_normalized_with_combat = data.frame(df_normalized_with_combat)
rownames(df_normalized_with_combat) = df_normalized_with_combat$cpgs # rownames for the csv
df_normalized_with_combat$cpgs = NULL

df_normalized$X205832330169_R03C01 = NULL #drop outlier
df_unnormalized$X205832330169_R03C01 = NULL
df_normalized_with_combat$X205832330169_R03C01 = NULL
df_masking_test$X205832330169_R03C01 = NULL

# load sample annotation
SampleAnnotation = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_SampleAnnotation_for_ibishof.Rds")
SampleAnnotation$grouping_variable = as.factor(SampleAnnotation$grouping_variable)
SampleAnnotation$XChipID

# adding age to SampleAnnotation
age = c()
chamber_number = c()
for (i in 1:length(SampleAnnotation$subjectLetter)){
  if (SampleAnnotation$subjectLetter[i]=="A") { age[i] = 24 }
  if (SampleAnnotation$subjectLetter[i]=="B") { age[i] = 39 }
  if (SampleAnnotation$subjectLetter[i]=="C") { age[i] = 54 }
  if (SampleAnnotation$subjectLetter[i]=="D") { age[i] = 66 }
}
SampleAnnotation$age = age

# Remove rows with nothing but NAs and median impute NA values
remove_na_median_impute <- function(dataframe){
  # Remove rows containing more than 10% NAs
  dataframe <- dataframe[rowSums(is.na(dataframe))/ncol(dataframe) < 0.1,]
  # Median impute remaining rows
  for (i in 1:nrow(dataframe)) {
    dataframe[i, which(is.na(dataframe[i,]))] <- median(as.numeric(dataframe[i,]),na.rm=TRUE)
  }
  class(as.numeric(dataframe[i,]))
  return(dataframe)
}

df_normalized = remove_na_median_impute(df_normalized)
df_unnormalized = remove_na_median_impute(df_unnormalized)
df_normalized_with_combat = remove_na_median_impute(df_normalized_with_combat)
#df_masking_test = remove_na_median_impute(df_masking_test)

```

From the methylclock vignette at bioconductor:

https://www.bioconductor.org/packages/release/bioc/vignettes/methylclock/inst/doc/methylclock.html#data-nomalization

"By default the package computes the different clocks when there are more than 80% of the required CpGs of each method. Nothing is required when having missing CpGs since the main functions will return NA for those estimators when this criteria is not meet [sic]."

```{r}


library(methylclock)

# Include cpgs in df_normalized, methylclock requires these be in the first column
cpgs = rownames(df_unnormalized)
df_unnormalized = cbind(cpgs,df_unnormalized)

cpgs = rownames(df_normalized)
df_normalized = cbind(cpgs,df_normalized)

cpgs = rownames(df_normalized)
df_normalized_with_combat = cbind(cpgs,df_normalized_with_combat)

cpgs = rownames(df_masking_test)
df_masking_test = cbind(cpgs,df_masking_test)

# check for missing cpgs
methylclock::checkClocks(df_unnormalized)
methylclock::checkClocks(df_normalized)
methylclock::checkClocks(df_normalized_with_combat)
methylclock::checkClocks(df_masking_test)

# calculate DNAmAge
age_unnormalized <- DNAmAge(df_unnormalized)
age_normalized <- DNAmAge(df_normalized)
#age_normalized_with_combat <- DNAmAge(df_normalized_with_combat)

# Values out of beta bounds after combat correction. Clipping here:
clip_dataframe <- function(df) {
  df[df < 0] <- 0
  df[df > 1] <- 1
  return(df)
}

df_normalized_with_combat = clip_dataframe(df_normalized_with_combat[,2:length(colnames(df_normalized_with_combat))])
cpgs = rownames(df_normalized)
df_normalized_with_combat = cbind(cpgs,df_normalized_with_combat)

age_normalized_with_combat <- DNAmAge(df_normalized_with_combat)

df_masking_test = clip_dataframe(df_masking_test[,2:length(colnames(df_masking_test))])
cpgs = rownames(df_masking_test)
df_masking_test = cbind(cpgs,df_masking_test)
age_normalized_with_combat_and_CV75 <- DNAmAge(df_masking_test)


```

```{r}
melted_age_normalized = reshape2::melt(age_normalized,ids = id)
melted_age_unnormalized = reshape2::melt(age_unnormalized,ids = id)
melted_normalized_with_combat = reshape2::melt(age_normalized_with_combat,ids = id)
melted_normalized_with_combat_and_CV75 = reshape2::melt(age_normalized_with_combat_and_CV75,ids = id)
```

Adding in subject letters
```{r}
subject = c()
chamber_number = c()
for (i in 1:length(melted_age_normalized$id)){
  index = which(SampleAnnotation$XChipID==melted_age_normalized$id[i])
  subject[i] = as.character(SampleAnnotation$subjectLetter[index])
  chamber_number[i] = as.character(SampleAnnotation$lane[index])
}
melted_age_normalized$subject = subject

chamber_number = c()
subject = c()
for (i in 1:length(melted_age_unnormalized$id)){
  index = which(SampleAnnotation$XChipID==melted_age_unnormalized$id[i])
  subject[i] = as.character(SampleAnnotation$subjectLetter[index])
  chamber_number[i] = as.character(SampleAnnotation$lane[index])
}
melted_age_unnormalized$subject = subject

chamber_number = c()
subject = c()
for (i in 1:length(melted_normalized_with_combat$id)){
  index = which(SampleAnnotation$XChipID==melted_normalized_with_combat$id[i])
  subject[i] = as.character(SampleAnnotation$subjectLetter[index])
  chamber_number[i] = as.character(SampleAnnotation$lane[index])
}
melted_normalized_with_combat$subject = subject

chamber_number = c()
subject = c()
for (i in 1:length(melted_normalized_with_combat_and_CV75$id)){
  index = which(SampleAnnotation$XChipID==melted_normalized_with_combat_and_CV75$id[i])
  subject[i] = as.character(SampleAnnotation$subjectLetter[index])
  chamber_number[i] = as.character(SampleAnnotation$lane[index])
}
melted_normalized_with_combat_and_CV75$subject = subject

# get rid of columns that aren't shared
keep = c("Horvath","Hannum","Levine","skinHorvath","PedBE","Wu","BLUP","EN") # manually identified
keep = c("Horvath","Hannum","Levine","skinHorvath","BLUP","EN") # manually identified

library(dplyr)
melted_normalized_with_combat = melted_normalized_with_combat %>%
  filter(variable %in% keep)
melted_age_unnormalized = melted_age_unnormalized %>%
  filter(variable %in% keep)
melted_age_normalized = melted_age_normalized %>%
  filter(variable %in% keep)
melted_normalized_with_combat_and_CV75 = melted_normalized_with_combat_and_CV75 %>%
  filter(variable %in% keep)
```

```{r}
saveRDS(melted_age_normalized,"melted_age_normalized.Rds")
saveRDS(melted_age_unnormalized,"melted_age_unnormalized.Rds")
saveRDS(melted_normalized_with_combat,"melted_normalized_with_combat.Rds")
saveRDS(melted_normalized_with_combat_and_CV75,"melted_normalized_with_combat_and_CV75.Rds")
```


Plots
```{r}


library(ggplot2)
graph2 <- ggplot(melted_age_normalized, aes(x = subject, y = value, colour = subject)) +
  facet_grid( . ~ variable ) +
  geom_point() +
  theme_bw() + 
  ggtitle("Age predictions pre-combat")
graph2

graph1 <- ggplot(melted_age_unnormalized, aes(x = subject, y = value, colour = subject)) +
  facet_grid( . ~ variable ) + 
  geom_point() +
  theme_bw() + 
  ggtitle("Age predictions raw")
graph1

graph3 <- ggplot(melted_normalized_with_combat, aes(x = subject, y = value, colour = subject)) +
  facet_grid( . ~ variable ) +
  geom_point() +
  theme_bw() + 
  ggtitle("Age predictions post-combat")
graph3

graph4 <- ggplot(melted_normalized_with_combat_and_CV75, aes(x = subject, y = value, colour = subject)) +
  facet_grid( . ~ variable ) +
  geom_point() +
  geom_hline(yintercept = c(24, 39, 54, 66), color = "red") + 
  theme_bw() + 
  ggtitle("Age predictions post-combat, CV filtered (<0.75)")
graph4

```

```{r}
age_graph_palette = paletteer::paletteer_d("trekcolors::lcars_2375", n = 4)

ages_raw <- ggplot(melted_age_unnormalized, aes(x = subject, y = value, colour = subject)) +
  paletteer::scale_color_paletteer_d("trekcolors::lcars_2375") + 
  facet_grid( . ~ variable ) +
  geom_point(size = 0.25, alpha=0.5) +
  geom_hline(yintercept = c(24), color = age_graph_palette[1],alpha=0.5) + 
  geom_hline(yintercept = c(39), color = age_graph_palette[2],alpha=0.5) + 
  geom_hline(yintercept = c(54), color = age_graph_palette[3],alpha=0.5) + 
  geom_hline(yintercept = c(66), color = age_graph_palette[4],alpha=0.5) + 
  theme_bw() + 
  ggtitle("Raw") +
  theme(
    axis.text.x = element_text(angle = 0, 
                               vjust = 1, 
                               hjust = 1),
    text = element_text(family = "Helvetica", 
                        size = 6), 
    legend.position = "none",
    axis.title = element_text(size = 6)
    )
ages_raw

ages_sesame <- ggplot(melted_age_normalized, aes(x = subject, y = value, colour = subject)) +
  paletteer::scale_color_paletteer_d("trekcolors::lcars_2375") + 
  facet_grid( . ~ variable ) +
  geom_point(size = 0.25, alpha=0.5) +
  geom_hline(yintercept = c(24), color = age_graph_palette[1],alpha=0.5) + 
  geom_hline(yintercept = c(39), color = age_graph_palette[2],alpha=0.5) + 
  geom_hline(yintercept = c(54), color = age_graph_palette[3],alpha=0.5) + 
  geom_hline(yintercept = c(66), color = age_graph_palette[4],alpha=0.5) + 
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 0, 
                               vjust = 1, 
                               hjust = 1),
    text = element_text(family = "Helvetica", 
                        size = 6),
    legend.position = "none",
    axis.title = element_text(size = 6)
    ) +
  ggtitle("SeSAMe")
ages_raw

ages_combat <- ggplot(melted_normalized_with_combat, aes(x = subject, y = value, colour = subject)) +
  paletteer::scale_color_paletteer_d("trekcolors::lcars_2375") + 
  facet_grid( . ~ variable ) +
  geom_point(size = 0.25, alpha=0.5) +
  geom_hline(yintercept = c(24), color = age_graph_palette[1],alpha=0.5) + 
  geom_hline(yintercept = c(39), color = age_graph_palette[2],alpha=0.5) + 
  geom_hline(yintercept = c(54), color = age_graph_palette[3],alpha=0.5) + 
  geom_hline(yintercept = c(66), color = age_graph_palette[4],alpha=0.5) + 
  theme_bw() + 
  theme(
    axis.text.x = element_text(angle = 0, 
                               vjust = 1, 
                               hjust = 1),
    text = element_text(family = "Helvetica", 
                        size = 6),
    legend.position = "none",
    axis.title = element_text(size = 6)
    ) +
  ggtitle("SeSAMe + ComBat")
ages_combat

```





```{r}
# # save_pdfs
# pdf("/home/sagemaker-user/dnamstability/dnamstability-abutler/pdf_plots/Fig7A_age_predictions_unnormalized.pdf",width = 6, height = 3,pointsize=10)
# graph1
# dev.off()
# 
# pdf("/home/sagemaker-user/dnamstability/dnamstability-abutler/pdf_plots/Fig7A_age_predictions_normalized.pdf",width = 6, height = 3,pointsize=10)
# graph2
# dev.off()
# 
# pdf("/home/sagemaker-user/dnamstability/dnamstability-abutler/pdf_plots/Fig7A_age_predictions_normalized_with_combat.pdf",width = 6, height = 3,pointsize=10)
# graph3
# dev.off()
```


# chamber number association with epigenetic age residual
```{r}
# wrangling columns
age_normalized = data.frame(age_normalized)
rownames(age_normalized) = age_normalized$id
age_normalized$id = NULL
age_normalized$BNN = NULL
age_normalized$TL = NULL

# calculate residuals
age_normalized_residuals = apply(age_normalized, 2, function(x) x - SampleAnnotation$age)
age_normalized_residuals = data.frame(age_normalized_residuals)
age_normalized_residuals$id = rownames(age_normalized)
melted_age_normalized_residuals = reshape2::melt(age_normalized_residuals,ids = id)

# adding chamber number
subject = c()
chamber_number = c()
for (i in 1:length(melted_age_normalized_residuals$id)){
  index = which(SampleAnnotation$XChipID==melted_age_normalized_residuals$id[i])
  subject[i] = as.character(SampleAnnotation$subjectLetter[index])
  chamber_number[i] = as.character(SampleAnnotation$lane[index])
}
melted_age_normalized_residuals$chamber_number = chamber_number
melted_age_normalized_residuals$subject = subject

graph <- ggplot(melted_age_normalized_residuals, aes(x = chamber_number, y = value, colour = chamber_number)) +
  geom_point(stat = 'summary', fun.y = 'mean') +
  geom_errorbar(stat = 'summary', fun.data = 'mean_se', 
                width=0, fun.args = list(mult = 1.96)) +
  facet_grid( subject ~ variable ) +
  labs(x = 'label x-axis', y = 'Age prediction residuals')+
  scale_colour_brewer(palette = 'Set2') +
  ggtitle("Chamber Number vs Age Prediction residuals") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


#age_unnormalized_residuals = apply(age_unnormalized, 2, function(x) x - SampleAnnotation$age)

graph = ggplot(melted_age_normalized_residuals, aes(x = chamber_number, y = value, colour = chamber_number)) +
  geom_point(stat = 'summary_bin',fun.data = 'mean_se') +
  geom_errorbar(stat = 'summary_bin', fun.data = 'mean_se', 
                width = 0, fun.args = list(mult = 1.96)) +
  facet_grid(subject ~ variable) +
  labs(x = 'label x-axis', y = 'Age prediction residuals') +
  scale_colour_brewer(palette = 'Set2') +
  ggtitle("Chamber Number vs Age Prediction residuals") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
graph

graph = ggplot(melted_age_normalized_residuals, aes(x = chamber_number, y = value, colour = chamber_number)) +
  geom_point(stat = 'summary_bin',fun.data = 'mean_se') +
  geom_errorbar(stat = 'summary_bin', fun.data = 'mean_se', 
                width = 0, fun.args = list(mult = 1.96)) +
  facet_grid(subject ~ variable) +
  labs(x = 'label x-axis', y = 'Age prediction residuals') +
  scale_colour_brewer(palette = 'Set2') +
  ggtitle("Chamber Number vs Age Prediction residuals") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
graph


# Preprocess the data to calculate mean and standard deviation
summary_data <- melted_age_normalized_residuals %>%
  group_by(chamber_number, variable) %>%
  summarize(mean_value = mean(value), sd_value = sd(value))


# Create the graph using the summary data
ggplotgui::ggplot_shiny(summary_data)
```

## looking at this a different way with linear mixed effect models
```{r}
library(lme4)
model <- lmer(value ~ variable + chamber_number + (1|subject), data = melted_age_normalized_residuals)
anova_results_temp = car::Anova(model, test = "F")
```

# chamber number vs age prediction & residual variability

```{r}
library(dplyr)
test = age_normalized %>%
  
```



