---
title: "Centered and scaled betas within subject - various comparisons"
output: html_notebook
---

The goal of this notebook is to center and scale beta values within subject in pooled samples, and make various comparisons.

A major goal is to compare unnormalized, normalized, and normalized + combat.

First, I will repeat the initial iteration of this analysis, where centered and scaled data were compared based on Sentrix Position (chamber number), with normalized, unnormalized, and normalized + combat data.

# Load data
```{r}
library(tidyverse)
library(knitr)
library(limma)
library(minfi)
library(RColorBrewer)
library(Gviz)
library(DMRcate)
library(stringr)
library(ggplot2)
library(ggfortify)
library(patchwork)
library(skimr)
library(esquisse)
library(rcartocolor)
library(paletteer)
library(dplyr)
library(readr)
```

```{r}

# load data
df_normalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221111_dnamstability_normalizedBetas_sesame_abutler.Rds")
df_unnormalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221111_dnamstability_unnormalizedBetas_sesame_abutler.Rds")
df_normalized_with_combat = read_csv("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_fluorescence_values/sesameAndCombat_normalized_betas_for_kras_20230317.csv")

rownames(df_normalized_with_combat) = df_normalized_with_combat$cpgs # rownames for the csv
df_normalized_with_combat$cpgs = NULL

df_normalized$X205832330169_R03C01 = NULL #drop outlier
df_unnormalized$X205832330169_R03C01 = NULL
df_normalized_with_combat$X205832330169_R03C01 = NULL


# load sample annotation
SampleAnnotation = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_SampleAnnotation_for_ibishof.Rds")
SampleAnnotation$grouping_variable = as.factor(SampleAnnotation$grouping_variable)
SampleAnnotation$XChipID


remove_na_10percent <- function(dataframe){
  dataframe <- dataframe[rowSums(is.na(dataframe))/ncol(dataframe) < 0.1,]
  return(dataframe)
}

df_normalized = remove_na_10percent(df_normalized)
df_unnormalized = remove_na_10percent(df_unnormalized)
df_normalized_with_combat = remove_na_10percent(df_normalized_with_combat)
```

# Standardize within-subject


```{r}
# Rewrote original code as a function
scale_data_by_grouping_variable <- function(data, annotation_df, subject_col = "XChipID", grouping_var = "subjectLetter") {
  groups <- unique(annotation_df[[grouping_var]])
  
  for (i in seq_along(groups)) {
  
    # select samples from each group (subjectLetter here)
    subject_df_temp <- data %>%
      select(any_of(annotation_df[[subject_col]][annotation_df[[grouping_var]]==groups[i]])) %>%
      select(any_of(annotation_df$XChipID[annotation_df$pooled==TRUE])) 
    
    # pause pipeline to save colnames (will be lost when using scale() below)
    colnames_temp = colnames(subject_df_temp) 
    
    # continue scaling
    subject_df_temp = subject_df_temp %>%
          apply(1,scale) %>%
          t() %>%
          as.data.frame()
    
    # reapply colnames
    colnames(subject_df_temp) = colnames_temp
    
    if (i == 1) {
      subject_df_scaled <- subject_df_temp
    } else {
      subject_df_scaled <- bind_cols(subject_df_temp,subject_df_scaled)
    }
  }
  return(subject_df_scaled)
}

# Applied
df_normalized_s = scale_data_by_grouping_variable(data = df_normalized, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter")
df_unnormalized_s = scale_data_by_grouping_variable(data = df_unnormalized, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter" )
df_normalized_with_combat_s = scale_data_by_grouping_variable(data = df_normalized_with_combat, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter" )

```

sanity check
```{r}
# each subject should now be centered
```





## Average across 4 subjects
```{r}
# pooled samples
df_unnormalized_s_pooled = df_unnormalized_s %>%
  dplyr::select(SampleAnnotation$XChipID[SampleAnnotation$pooled==TRUE])
df_normalized_s_pooled = df_normalized_s %>%
  dplyr::select(SampleAnnotation$XChipID[SampleAnnotation$pooled==TRUE])
df_normalized_with_combat_s_pooled = df_normalized_with_combat_s %>%
  dplyr::select(SampleAnnotation$XChipID[SampleAnnotation$pooled==TRUE])

# average_by_subjects_in_pooled_samples
average_by_subjects_in_pooled_samples = function(df){
positions = unique(SampleAnnotation$lane)
for (i in 1:length(positions)){
  temp = df %>%
    dplyr::select(SampleAnnotation$XChipID[SampleAnnotation$pooled==TRUE]) %>%
    dplyr::select(any_of(SampleAnnotation$XChipID[SampleAnnotation$lane==positions[i]])) %>% apply(1,mean,na.rm=TRUE)
  if (i == 1){
    temp2 = data.frame(CpG = rownames(df),
                       temp1 = temp)
  } else {
    temp2 = cbind(temp2, temp)
  }
}
  return(temp2)
}

# chamber number means across four subjects
df_unnormalized_s_pooled_means = average_by_subjects_in_pooled_samples(df_unnormalized_s_pooled)

df_normalized_s_pooled_means = average_by_subjects_in_pooled_samples(df_normalized_s_pooled)

df_normalized_with_combat_s_pooled_means = average_by_subjects_in_pooled_samples(df_normalized_with_combat_s_pooled)

```


Melt columns and add in ID variables
```{r}
ids_to_bind = t(SampleAnnotation)
ids_to_bind = data.frame(ids_to_bind)
colnames(ids_to_bind) = ids_to_bind["XChipID",]

# putting variable info into a single string string (new column name) to preserve while melting
SampleAnnotation$BigName = paste(SampleAnnotation$XChipID, 
                                 SampleAnnotation$subjectLetter,
                                 SampleAnnotation$pooled,
                                 SampleAnnotation$grouping_variable_bswa,
                                 SampleAnnotation$grouping_variable,
                                 SampleAnnotation$grouping_variable2,
                                 SampleAnnotation$array,
                                 SampleAnnotation$lane,
                                 sep = '.')

for (i in 1:length(colnames(df_normalized_with_combat_s))){
  j = which(SampleAnnotation$XChipID==colnames(df_normalized_with_combat_s)[i])
  colnames(df_normalized_with_combat_s)[i] = SampleAnnotation$BigName[j]
}
df_normalized_with_combat_s = reshape2::melt(df_normalized_with_combat_s)

for (i in 1:length(colnames(df_normalized_s))){
  j = which(SampleAnnotation$XChipID==colnames(df_normalized_s)[i])
  colnames(df_normalized_s)[i] = SampleAnnotation$BigName[j]
}
df_normalized_s = reshape2::melt(df_normalized_s)

for (i in 1:length(colnames(df_unnormalized_s))){
  j = which(SampleAnnotation$XChipID==colnames(df_unnormalized_s)[i])
  colnames(df_unnormalized_s)[i] = SampleAnnotation$BigName[j]
}
df_unnormalized_s = reshape2::melt(df_unnormalized_s)

# Split the variable info back out (delimited by ".")
split_column_name <- function(df, column_name, delimiter) {
  col_data <- as.character(df[[column_name]]) # convert column to character type
  split_data <- strsplit(col_data, delimiter) # split the data by the delimiter
  split_data <- do.call(rbind, split_data) # convert the list to a matrix
  df <- cbind(df, split_data) # append split data as new columns
  return(df)
}

df_normalized_s_pooled_means = split_column_name(df_normalized_s,"variable","\\.")
df_unnormalized_s_pooled_means = split_column_name(df_unnormalized_s,"variable","\\.")
df_normalized_with_combat_s_pooled_means = split_column_name(df_normalized_with_combat_s,"variable","\\.")



```

```{r}
# Rename columns
new_colnames = c("variable","value","XChipID","subject","pooled","grouping_var_bswa","grouping_var","grouping_var2","array","lane")
colnames(df_normalized_s_pooled_means) = new_colnames
colnames(df_normalized_with_combat_s_pooled_means) = new_colnames
colnames(df_unnormalized_s_pooled_means) = new_colnames
```

Make plots
```{r}
graph1 <- ggplot(df_unnormalized_s_pooled_means, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05, show.legend = FALSE) +
  theme_bw() + 
  facet_grid( lane ~ . ) +
  ggtitle("Distribution of centered and scaled beta values in unnormalized data",
          "Avg of 4 subjects")
graph1

graph2<- ggplot(df_normalized_s_pooled_means, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05, show.legend = FALSE) +
  theme_bw() +
  facet_grid( lane ~ . ) +
  ggtitle("Distribution of centered and scaled beta values in normalized data",
          "Avg of 4 subjects")
graph2

graph3 <- ggplot(df_normalized_with_combat_s_pooled_means, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05, show.legend = FALSE) +
  theme_bw() +
  facet_grid( lane ~ . )+
  ggtitle("Distribution of centered and scaled beta values in normalized data",
          "Avg of 4 subjects - Sentrix position corrected using ComBat")
graph3

# same as last figure, but with the legend there so we can crop this
graph4 <- ggplot(df_normalized_with_combat_s_pooled_means, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05) +
  theme_bw() +
  facet_grid( lane ~ . ) +
  ggtitle("Distribution of centered and scaled beta values in normalized data",
          "Avg of 4 subjects - Sentrix position corrected using ComBat")
graph4

composite_graph = graph1 + graph2 + graph3
composite_graph
```

### Fig3A_Distributions_of_centered_and_scaled_betas.pdf
```{r, Fig3A_Distributions_of_centered_and_scaled_betas.pdf}
# save_pdf
pdf("/home/sagemaker-user/dnamstability/dnamstability-abutler/pdf_plots/Fig3A_Distributions_of_centered_and_scaled_betas.pdf",width = 5, height = 3,pointsize=10)
composite_graph
dev.off()
```


## Average across 4 arrays
```{r}
# average_by_subjects_in_pooled_samples
average_arrays_instead_of_chamber_number = function(df){
arrays = unique(SampleAnnotation$array[SampleAnnotation$pooled==TRUE])
for (i in 1:length(arrays)){
  temp = df %>%
    dplyr::select(SampleAnnotation$XChipID[SampleAnnotation$pooled==TRUE]) %>%
    dplyr::select(any_of(SampleAnnotation$XChipID[SampleAnnotation$array==arrays[i]])) %>% apply(1,mean,na.rm=TRUE)
  if (i == 1){
    temp2 = data.frame(CpG = rownames(df),
                       temp1 = temp)
  } else {
    temp2 = cbind(temp2, temp)
  }
}
  return(temp2)
}
df_unnormalized_s_pooled_arrayMeans = average_arrays_instead_of_chamber_number(df_unnormalized_s_pooled)

df_normalized_s_pooled_arrayMeans = average_arrays_instead_of_chamber_number(df_normalized_s_pooled)

df_normalized_with_combat_s_pooled_arrayMeans = average_arrays_instead_of_chamber_number(df_normalized_with_combat_s_pooled)

arrays = unique(SampleAnnotation$array[SampleAnnotation$pooled==TRUE])
array_colnames = c("CpG",as.character(arrays))

colnames(df_unnormalized_s_pooled_arrayMeans) =array_colnames
colnames(df_normalized_s_pooled_arrayMeans) =array_colnames
colnames(df_normalized_with_combat_s_pooled_arrayMeans) =array_colnames

```

```{r}
melted_df_unnormalized_s_pooled_arrayMeans = reshape2::melt(df_unnormalized_s_pooled_arrayMeans,id.vars = "CpG")

melted_df_normalized_s_pooled_arrayMeans = reshape2::melt(df_normalized_s_pooled_arrayMeans,id.vars = "CpG")

melted_df_normalized_with_combat_s_pooled_arrayMeans = reshape2::melt(df_normalized_with_combat_s_pooled_arrayMeans,id.vars = "CpG")




# The code below will generate the graph:
graph1 <- ggplot(melted_df_unnormalized_s_pooled_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02,show.legend = FALSE) +
  theme_bw() + 
    scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph1

graph2 <- ggplot(melted_df_normalized_s_pooled_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02,show.legend = FALSE) +
  theme_bw() + 
  scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph2

graph3 <- ggplot(melted_df_normalized_with_combat_s_pooled_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02,show.legend = FALSE) +
  theme_bw() + 
  scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph3

composite_graph = graph1 + graph2 + graph3

graph4 <- ggplot(melted_df_normalized_with_combat_s_pooled_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02) +
  theme_bw() + 
  scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph4
```

```{r, Fig3B_Distributions_of_centered_and_scaled_betas_arrays.pdf}
# save_pdf
pdf("/home/sagemaker-user/dnamstability/dnamstability-abutler/pdf_plots/Fig3A_Distributions_of_centered_and_scaled_betas_in_arrays.pdf",width = 5, height = 3,pointsize=10)
composite_graph
dev.off()

```

# Average across arrays not just pooled

```{r}
scale_data_by_grouping_variable <- function(data, annotation_df, subject_col = "XChipID", grouping_var = "subjectLetter") {
  groups <- unique(annotation_df[[grouping_var]])
  
  for (i in seq_along(groups)) {
  
    # select samples from each group (subjectLetter here)
    subject_df_temp <- data %>%
      select(any_of(annotation_df[[subject_col]][annotation_df[[grouping_var]]==groups[i]])) #%>%
      #select(any_of(annotation_df$XChipID[annotation_df$pooled==TRUE])) 
    
    # pause pipeline to save colnames (will be lost when using scale() below)
    colnames_temp = colnames(subject_df_temp) 
    
    # continue scaling
    subject_df_temp = subject_df_temp %>%
          apply(1,scale) %>%
          t() %>%
          as.data.frame()
    
    # reapply colnames
    colnames(subject_df_temp) = colnames_temp
    
    if (i == 1) {
      subject_df_scaled <- subject_df_temp
    } else {
      subject_df_scaled <- bind_cols(subject_df_temp,subject_df_scaled)
    }
  }
  return(subject_df_scaled)
}

# Applied
df_normalized_s_pooledAndNumbered = scale_data_by_grouping_variable(data = df_normalized, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter")
df_unnormalized_s_pooledAndNumbered = scale_data_by_grouping_variable(data = df_unnormalized, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter" )
df_normalized_with_combat_s_pooledAndNumbered = scale_data_by_grouping_variable(data = df_normalized_with_combat, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter" )

# average_by_subjects
average_arrays_instead_of_chamber_number = function(df){
arrays = unique(SampleAnnotation$array)
for (i in 1:length(arrays)){
  temp = df %>%
    dplyr::select(any_of(SampleAnnotation$XChipID[SampleAnnotation$array==arrays[i]])) %>% apply(1,mean,na.rm=TRUE)
  if (i == 1){
    temp2 = data.frame(CpG = rownames(df),
                       temp1 = temp)
  } else {
    temp2 = cbind(temp2, temp)
  }
}
  return(temp2)
}
df_unnormalized_s_pooledAndNumbered_arrayMeans = average_arrays_instead_of_chamber_number(df_unnormalized_s_pooledAndNumbered)

df_normalized_s_pooledAndNumbered_arrayMeans = average_arrays_instead_of_chamber_number(df_normalized_s_pooledAndNumbered)

df_normalized_with_combat_s_pooledAndNumbered_arrayMeans = average_arrays_instead_of_chamber_number(df_normalized_with_combat_s_pooledAndNumbered)

arrays = unique(SampleAnnotation$array)
array_colnames = c("CpG",as.character(arrays))

colnames(df_unnormalized_s_pooledAndNumbered_arrayMeans) =array_colnames
colnames(df_normalized_s_pooledAndNumbered_arrayMeans) =array_colnames
colnames(df_normalized_with_combat_s_pooledAndNumbered_arrayMeans) =array_colnames


melted_df_unnormalized_s_pooledAndNumbered_arrayMeans = reshape2::melt(df_unnormalized_s_pooledAndNumbered_arrayMeans,id.vars = "CpG")

melted_df_normalized_s_pooledAndNumbered_arrayMeans = reshape2::melt(df_normalized_s_pooledAndNumbered_arrayMeans,id.vars = "CpG")

melted_df_normalized_with_combat_s_pooledAndNumbered_arrayMeans = reshape2::melt(df_normalized_with_combat_s_pooledAndNumbered_arrayMeans,id.vars = "CpG")




# The code below will generate the graph:
graph1 <- ggplot(melted_df_unnormalized_s_pooledAndNumbered_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02,show.legend = FALSE) +
  theme_bw() + 
    scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph1

graph2 <- ggplot(melted_df_normalized_s_pooledAndNumbered_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02,show.legend = FALSE) +
  theme_bw() + 
  scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph2

graph3 <- ggplot(melted_df_normalized_with_combat_s_pooledAndNumbered_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02,show.legend = FALSE) +
  theme_bw() + 
  scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph3

composite_graph = graph1 + graph2 + graph3

graph4 <- ggplot(melted_df_normalized_with_combat_s_pooledAndNumbered_arrayMeans, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.02) +
  theme_bw() + 
  scale_fill_paletteer_d("colorblindr::OkabeIto") + 
  facet_grid(rows = "variable")
graph4

composite_graph_means = graph1 + graph2 + graph3
composite_graph_means
```






## curves before taking averages

```{r}
df_unnormalized_s_means_pooled = df_unnormalized_s %>%
  filter(pooled = TRUE)
df_normalized_s_means_pooled = df_normalized_s %>%
  filter(pooled = TRUE)
df_normalized_with_combat_s_means_pooled = df_normalized_with_combat_s %>%
  filter(pooled = TRUE)

df_unnormalized_s_means_pooled = apply(df_unnormalized_s_means_pooled,1,mean,na.rm=TRUE )
df_normalized_s_means_pooled = apply(df_normalized_s_means_pooled,1,mean,na.rm=TRUE )
df_normalized_with_combat_s_means_pooled = apply(df_normalized_with_combat_s_means_pooled,1,mean,na.rm=TRUE )



graph1 <- ggplot(df_unnormalized_s_means_pooled, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05) +
  theme_bw() + 
  facet_grid( lane ~ . ) + 
  ggtitle("Distribution of centered and scaled beta values in unnormalized data",
          "Avg of 4 subjects")
graph1

graph2<- ggplot(df_normalized_s_means_pooled, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05) +
  theme_bw() +
  facet_grid( lane ~ . ) + 
  ggtitle("Distribution of centered and scaled beta values in normalized data",
          "Avg of 4 subjects")
graph2

graph3 <- ggplot(df_normalized_with_combat_s_means_pooled, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05) +
  theme_bw() +
  facet_grid( lane ~ . ) +
  ggtitle("Distribution of centered and scaled beta values in normalized data",
          "Avg of 4 subjects - Sentrix position corrected using ComBat")
graph3
```


## curves without taking averages
```{r}
#ggplotgui::ggplot_shiny(df_unnormalized_s)
df_unnormalized_s_pooled = df_unnormalized_s %>%
  filter(pooled = TRUE)
df_normalized_s_pooled = df_normalized_s %>%
  filter(pooled = TRUE)
df_normalized_with_combat_s_pooled = df_normalized_with_combat_s %>%
  filter(pooled = TRUE)



# The code below will generate the graph:
graph1 <- ggplot(df_unnormalized_s_pooled, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05) +
  theme_bw() + 
  facet_grid( lane ~ . ) + 
  ggtitle("Distribution of centered and scaled beta values in unnormalized data")
graph1

graph2<- ggplot(df_normalized_s_pooled, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05) +
  theme_bw() +
  facet_grid( lane ~ . ) + 
  ggtitle("Distribution of centered and scaled beta values in normalized data")
graph2

graph3 <- ggplot(df_normalized_with_combat_s_pooled, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05) +
  theme_bw() +
  facet_grid( lane ~ . ) +
  ggtitle("Distribution of centered and scaled beta values in normalized data",
          "Sentrix position corrected using ComBat")
graph3


```






```{r}
graph1 <- ggplot(df_unnormalized_s, aes(x = value, fill = array)) +
  geom_histogram(position = 'identity', alpha = 0.5, binwidth = 0.05) +
  theme_bw() + 
  facet_grid( array ~ . )
graph

graph2 <- ggplot(df_unnormalized_s, aes(x = value, fill = array)) +
  geom_histogram(position = 'identity', alpha = 0.5, binwidth = 0.05) +
  theme_bw() +
  facet_grid( array ~ . )
graph

graph3 <- ggplot(df_normalized_with_combat_s, aes(x = value, fill = array)) +
  geom_histogram(position = 'identity', alpha = 0.5, binwidth = 0.05) +
  theme_bw() +
  facet_grid( array ~ . )
graph
```


```{r}
graph <- ggplot(df_unnormalized_s, aes(x = value, fill = array)) +
  geom_histogram(position = 'identity', alpha = 0.5, binwidth = 0.05) +
  theme_bw() + 
  facet_grid( array ~ . )
graph

graph <- ggplot(df_unnormalized_s, aes(x = value, fill = array)) +
  geom_histogram(position = 'identity', alpha = 0.5, binwidth = 0.05) +
  theme_bw() +
  facet_grid( array ~ . )
graph

graph <- ggplot(df_normalized_with_combat_s, aes(x = value, fill = array)) +
  geom_histogram(position = 'identity', alpha = 0.5, binwidth = 0.05) +
  theme_bw() +
  facet_grid( array ~ . )
graph
```

```{r}

```

# FI - load data
```{r}
# raw color values
unmethylated_raw = read_csv("data/raw_fluorescence_values/unmethylated_raw.csv")
unmethylated_raw = data.frame(unmethylated_raw)
rownames(unmethylated_raw) = unmethylated_raw$...1
unmethylated_raw$...1 = NULL

methylated_raw = read_csv("data/raw_fluorescence_values/methylated_raw.csv")
methylated_raw = data.frame(methylated_raw)
rownames(methylated_raw) = methylated_raw$...1
methylated_raw$...1 = NULL

sesame_jkras_type2 = read_csv("sesame_jkras_type2_20230317.csv")

methylated_raw$cpgs = rownames(methylated_raw)
methylated_raw = methylated_raw %>%
  filter(cpgs %in% sesame_jkras_type2$cpgs)
methylated_raw$cpgs = NULL

unmethylated_raw$cpgs = rownames(unmethylated_raw)
unmethylated_raw = unmethylated_raw %>%
  filter(cpgs %in% sesame_jkras_type2$cpgs)
unmethylated_raw$cpgs = NULL

# load sample annotation
SampleAnnotation = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_SampleAnnotation_for_ibishof.Rds")
SampleAnnotation$grouping_variable = as.factor(SampleAnnotation$grouping_variable)
SampleAnnotation$XChipID
```

# FI - methylated

# Standardize within-subject

```{r}
# Rewrote original code as a function
scale_data_by_grouping_variable <- function(data, annotation_df, subject_col = "XChipID", grouping_var = "subjectLetter") {
  groups <- unique(annotation_df[[grouping_var]])
  
  for (i in seq_along(groups)) {
  
    # select samples from each group (subjectLetter here)
    subject_df_temp <- data %>%
      select(any_of(annotation_df[[subject_col]][annotation_df[[grouping_var]]==groups[i]])) %>%
      select(any_of(annotation_df$XChipID[annotation_df$pooled==TRUE])) 
    
    # pause pipeline to save colnames (will be lost when using scale() below)
    colnames_temp = colnames(subject_df_temp) 
    
    # continue scaling
    subject_df_temp = subject_df_temp %>%
          apply(1,scale) %>%
          t() %>%
          as.data.frame()
    
    # reapply colnames
    colnames(subject_df_temp) = colnames_temp
    
    if (i == 1) {
      subject_df_scaled <- subject_df_temp
    } else {
      subject_df_scaled <- bind_cols(subject_df_temp,subject_df_scaled)
    }
  }
  return(subject_df_scaled)
}

# Applied
methylated_raw_s = scale_data_by_grouping_variable(data = methylated_raw, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter")
unmethylated_raw_s = scale_data_by_grouping_variable(data = unmethylated_raw, 
                                                  annotation_df = SampleAnnotation,
                                                  subject_col = "XChipID", 
                                                  grouping_var = "subjectLetter" )
```


## Average across 4 subjects
```{r}

# average_by_subjects
average_by_subjects_in_pooled_samples = function(df){
positions = unique(SampleAnnotation$lane)
for (i in 1:length(positions)){
  temp = df %>%
    dplyr::select(SampleAnnotation$XChipID[SampleAnnotation$pooled==TRUE]) %>%
    dplyr::select(any_of(SampleAnnotation$XChipID[SampleAnnotation$lane==positions[i]])) %>% apply(1,mean,na.rm=TRUE)
  if (i == 1){
    temp2 = data.frame(CpG = rownames(df),
                       temp1 = temp)
  } else {
    temp2 = cbind(temp2, temp)
  }
}
  return(temp2)
}

# chamber number means across four subjects
unmethylated_raw_s_pooled_means = average_by_subjects_in_pooled_samples(unmethylated_raw_s)

methylated_raw_s_pooled_means = average_by_subjects_in_pooled_samples(methylated_raw_s)

colnames(unmethylated_raw_s_pooled_means) = c("CpG","1","2","3","4","5","6","7","8")
rownames(unmethylated_raw_s_pooled_means) = unmethylated_raw_s_pooled_means$CpG
unmethylated_raw_s_pooled_means$CpG = NULL
unmethylated_raw_s_pooled_means = reshape2::melt(unmethylated_raw_s_pooled_means)

colnames(methylated_raw_s_pooled_means) = c("CpG","1","2","3","4","5","6","7","8")
rownames(methylated_raw_s_pooled_means) = methylated_raw_s_pooled_means$CpG
methylated_raw_s_pooled_means$CpG = NULL
methylated_raw_s_pooled_means = reshape2::melt(methylated_raw_s_pooled_means)

```

# Make plots - means
```{r}
library(monochromeR)

reds = generate_palette("red", modification = "go_darker", 
                 n_colours = 8, view_palette = TRUE)

greens = generate_palette("green", modification = "go_darker", 
                 n_colours = 8, view_palette = TRUE)


graph1 <- ggplot(unmethylated_raw_s_pooled_means, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05, show.legend = FALSE) +
  theme_bw() + 
  facet_grid( variable ~ . ) +
  #ylim(0,175000) +
  ggtitle("Red (unmethylated)")+
  scale_fill_manual(values = reds)+
  theme(text = element_text(size = 5)) 
graph1

graph2<- ggplot(methylated_raw_s_pooled_means, aes(x = value, fill = variable)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05, show.legend = FALSE) +
  theme_bw() +
  facet_grid( variable ~ . ) +
  #ylim(0,175000) +
  ggtitle("Green (methylated)")+
  scale_fill_manual(values = greens)+
  theme(text = element_text(size = 5)) 
graph2

composite_graph = graph1 + graph2
composite_graph
```

```{r}
### Plotting
pdf_export_directory = '~/dnamstability/dnamstability-abutler/pdf_plots/' #where I'm saving figures
pdf(file = paste0(pdf_export_directory, "distributions_of_centered_and_scaled_FI_means.pdf"),   # The directory you want to save the file in
    width = (90/25.4), # The width of the plot in inches (converted from mm)
    height = (90/25.4)) # The height of the plot in inches
composite_graph
dev.off()
###
```




Melt columns and add in ID variables
```{r}
ids_to_bind = t(SampleAnnotation)
ids_to_bind = data.frame(ids_to_bind)
colnames(ids_to_bind) = ids_to_bind["XChipID",]

# putting variable info into a single string string (new column name) to preserve while melting
SampleAnnotation$BigName = paste(SampleAnnotation$XChipID, 
                                 SampleAnnotation$subjectLetter,
                                 SampleAnnotation$pooled,
                                 SampleAnnotation$grouping_variable_bswa,
                                 SampleAnnotation$grouping_variable,
                                 SampleAnnotation$grouping_variable2,
                                 SampleAnnotation$array,
                                 SampleAnnotation$lane,
                                 sep = '.')

for (i in 1:length(colnames(methylated_raw_s))){
  j = which(SampleAnnotation$XChipID==colnames(methylated_raw_s)[i])
  colnames(methylated_raw_s)[i] = SampleAnnotation$BigName[j]
}
methylated_raw_s = reshape2::melt(methylated_raw_s)

for (i in 1:length(colnames(unmethylated_raw_s))){
  j = which(SampleAnnotation$XChipID==colnames(unmethylated_raw_s)[i])
  colnames(unmethylated_raw_s)[i] = SampleAnnotation$BigName[j]
}
unmethylated_raw_s = reshape2::melt(unmethylated_raw_s)

# Split the variable info back out (delimited by ".")
split_column_name <- function(df, column_name, delimiter) {
  col_data <- as.character(df[[column_name]]) # convert column to character type
  split_data <- strsplit(col_data, delimiter) # split the data by the delimiter
  split_data <- do.call(rbind, split_data) # convert the list to a matrix
  df <- cbind(df, split_data) # append split data as new columns
  return(df)
}

methylated_raw_s_pooled_means = split_column_name(methylated_raw_s,"variable","\\.")
unmethylated_raw_s_pooled_means = split_column_name(unmethylated_raw_s,"variable","\\.")
```

```{r}
# Rename columns
new_colnames = c("variable","value","XChipID","subject","pooled","grouping_var_bswa","grouping_var","grouping_var2","array","lane")
colnames(methylated_raw_s_pooled_means) = new_colnames
colnames(unmethylated_raw_s_pooled_means) = new_colnames
```

# Make plots
```{r}
library(monochromeR)

reds = generate_palette("red", modification = "go_darker", 
                 n_colours = 8, view_palette = TRUE)

greens = generate_palette("green", modification = "go_darker", 
                 n_colours = 8, view_palette = TRUE)


graph1 <- ggplot(unmethylated_raw_s_pooled_means, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05, show.legend = FALSE) +
  theme_bw() + 
  facet_grid( lane ~ . ) +
  #ylim(0,175000) +
  ggtitle("Red (unmethylated)")+
  scale_fill_manual(values = reds)+
  theme(text = element_text(size = 5)) 
graph1

graph2<- ggplot(methylated_raw_s_pooled_means, aes(x = value, fill = lane)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.05, show.legend = FALSE) +
  theme_bw() +
  facet_grid( lane ~ . ) +
  #ylim(0,175000) +
  ggtitle("Green (methylated)")+
  scale_fill_manual(values = greens)+
  theme(text = element_text(size = 5)) 
graph2

composite_graph = graph1 + graph2
composite_graph
```
```{r}
### Plotting
pdf_export_directory = '~/dnamstability/dnamstability-abutler/pdf_plots/' #where I'm saving figures
pdf(file = paste0(pdf_export_directory, "distributions_of_centered_and_scaled_FIs.pdf"),   # The directory you want to save the file in
    width = (90/25.4), # The width of the plot in inches (converted from mm)
    height = (90/25.4)) # The height of the plot in inches
composite_graph
dev.off()
###
```


```{r}
### Plotting
pdf_export_directory = '~/dnamstability/dnamstability-abutler/pdf_plots/' #where I'm saving figures
pdf(file = paste0(pdf_export_directory, "distributions_of_centered_and_scaled_FI_means.pdf"),   # The directory you want to save the file in
    width = (90/25.4), # The width of the plot in inches (converted from mm)
    height = (90/25.4)) # The height of the plot in inches
composite_graph_means
dev.off()
###
```