---
title: "R Notebook"
output: html_notebook
---

# Exploration of a single variable CpG from the Skin clock to see what's going on with this CpG

## cg03183882: A single CpG with high range and high coefficient in the skin clock

check the probe type
```{r}
library(ggplot2)
library(readr)
library(dplyr)
library(minfi)

# check the probe type:
infinium_methylationepic_v_1_0_b5_manifest_file <- read_csv("infinium-methylationepic-v-1-0-b5-manifest-file.csv",
    skip = 7)

cg03183882 = infinium_methylationepic_v_1_0_b5_manifest_file[infinium_methylationepic_v_1_0_b5_manifest_file$IlmnID=="cg03183882",] # it's a typeII probe
```
## load data
```{r}
# Load raw data
df_normalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_normalizedBetas_raw_abutler.Rds")

df_unnormalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_unnormalizedBetas_raw_abutler.Rds")

SampleAnnotation = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_SampleAnnotation_for_ibishof.Rds")

df_normalized$X205832330169_R03C01 = NULL
df_unnormalized$X205832330169_R03C01 = NULL

SampleAnnotation$row_number = substr(SampleAnnotation$ChIP_ID,16,16)

SampleAnnotation = SampleAnnotation %>%
  dplyr::select(ChIP_ID,subject,row_number,array,pooled)
SampleAnnotation$XChIP_ID = paste0("X",SampleAnnotation$ChIP_ID)
#SampleAnnotation$ChIP_ID=NULL

# make rgSet extended
idat_dir = file.path('/home/sagemaker-user/dnamstability/dnamstability-abutler/data/idat_files/')
rgSet = read.metharray.exp(idat_dir, extended = TRUE)
```


```{r}
# get low-level data for probes
TypeI.Red <- getProbeInfo(rgSet, type = "I-Red")
TypeI.Green <- getProbeInfo(rgSet, type = "I-Green")
TypeII <- getProbeInfo(rgSet, type = "II")

Grn = assay(rgSet, "Green")       # Green mean across beads
Red = assay(rgSet, "Red")         # Red mean across beads
GrnSD = assay(rgSet, "GreenSD")   # Green SD across beads
RedSD = assay(rgSet, "RedSD")     # Red SD across beads
nBeads = assay(rgSet, "NBeads")   # Number of Beads across probes

df_RedSD = as.data.frame(RedSD)
df_GrnSD = as.data.frame(GrnSD)
df_Red = as.data.frame(Red)
df_Green = as.data.frame(Grn)
df_nBeads = as.data.frame(nBeads)

typeII_red_stdev = df_RedSD %>%
  filter((rownames(RedSD) %in% TypeII$AddressA))

typeII_green_stdev = df_GrnSD %>%
  filter((rownames(GrnSD) %in% TypeII$AddressA))

typeII_red = df_Red %>%
  filter((rownames(Red) %in% TypeII$AddressA))

typeII_green = df_Green %>%
  filter((rownames(Grn) %in% TypeII$AddressA))

typeII_nBeads = df_nBeads %>%
  filter((rownames(nBeads) %in% TypeII$AddressA))

# add cpg name info
TypeII_key = data.frame(Name = TypeII$Name,
                        AddressA = TypeII$AddressA)

typeII_red_stdev = merge(typeII_red_stdev,TypeII_key,by.x="row.names",by.y = "AddressA")
typeII_green_stdev = merge(typeII_green_stdev,TypeII_key,by.x="row.names",by.y = "AddressA")
typeII_red = merge(typeII_red,TypeII_key,by.x="row.names",by.y = "AddressA")
typeII_green = merge(typeII_green,TypeII_key,by.x="row.names",by.y = "AddressA")
typeII_nBeads = merge(typeII_nBeads,TypeII_key,by.x="row.names",by.y = "AddressA")
```

```{r}
# filter for cg03183882
cg03183882_red_stdev = typeII_red_stdev[typeII_red_stdev$Name=="cg03183882",]
cg03183882_green_stdev = typeII_green_stdev[typeII_green_stdev$Name=="cg03183882",]
cg03183882_red = typeII_red[typeII_red$Name=="cg03183882",]
cg03183882_green = typeII_green[typeII_green$Name=="cg03183882",]
cg03183882_nBeads = typeII_nBeads[typeII_nBeads$Name=="cg03183882",]


# filter columns
cg03183882_red_stdev = cg03183882_red_stdev[,colnames(cg03183882_red_stdev) %in% SampleAnnotation$ChIP_ID]
cg03183882_red_stdev = reshape2::melt(cg03183882_red_stdev)
cg03183882_red_stdev$feature = "red_stdev"
cg03183882_green_stdev = cg03183882_green_stdev[,colnames(cg03183882_green_stdev) %in% SampleAnnotation$ChIP_ID]
cg03183882_green_stdev = reshape2::melt(cg03183882_green_stdev)
cg03183882_green_stdev$feature = "green_stdev"
cg03183882_red = cg03183882_red[,colnames(cg03183882_red) %in% SampleAnnotation$ChIP_ID]
cg03183882_red = reshape2::melt(cg03183882_red)
cg03183882_red$feature = "red"

cg03183882_green = cg03183882_green[,colnames(cg03183882_green) %in% SampleAnnotation$ChIP_ID]
cg03183882_green = reshape2::melt(cg03183882_green)
cg03183882_green$feature = "green"
cg03183882_nBeads = cg03183882_nBeads[,colnames(cg03183882_nBeads) %in% SampleAnnotation$ChIP_ID]
cg03183882_nBeads = reshape2::melt(cg03183882_nBeads)
cg03183882_nBeads$feature = "nBeads"

cg03183882_lowlevel = rbind(cg03183882_nBeads,
                            cg03183882_red,
                            cg03183882_green,
                            cg03183882_red_stdev,
                            cg03183882_green_stdev)
cg03183882_lowlevel_cols = cbind(cg03183882_nBeads,
                            cg03183882_red,
                            cg03183882_green,
                            cg03183882_red_stdev,
                            cg03183882_green_stdev)




cg03183882_lowlevel$subject = "E"
for (i in 1:length(cg03183882_lowlevel$variable)){
  j = which(SampleAnnotation$ChIP_ID == cg03183882_lowlevel$variable[i])
  cg03183882_lowlevel$subject[i] = SampleAnnotation$subject[j]
}

cg03183882_lowlevel_fluorescence = cg03183882_lowlevel
cg03183882_lowlevel_fluorescence = cg03183882_lowlevel_fluorescence[cg03183882_lowlevel_fluorescence$feature!="nBeads",]
ggplotgui::ggplot_shiny(cg03183882_lowlevel_fluorescence)




# The code below will generate graphs; however, screenshots of plotly plots were taken using ggplotgui
graph_for_fig <- ggplot(cg03183882_lowlevel_fluorescence, aes(x = variable, y = value, colour = feature)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1),text = element_text(family = "Helvetica", size = 6), legend.position = "none",
    axis.title = element_text(size = 6)) + 
  scale_color_manual(values = c("green","darkgreen","red","darkred"))+
  ggtitle("cg03183882 - Raw")

graph_for_fig

graph1_withLegend <- ggplot(cg03183882_lowlevel_fluorescence, aes(x = variable, y = value, colour = feature)) +
  geom_point() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1),text = element_text(family = "Helvetica", size = 6), 
    axis.title = element_text(size = 6)) + 
  scale_color_manual(values = c("green","darkgreen","red","darkred"))+
  ggtitle("cg03183882 - Raw")
graph1_withLegend

graph2 <- ggplot(cg03183882_nBeads, aes(x = variable, y = value)) +
  geom_point() +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
graph2

outliers::grubbs.test(cg03183882_green_stdev$value)
outliers::grubbs.test(cg03183882_red_stdev$value)
outliers::grubbs.test(cg03183882_green$value)
outliers::grubbs.test(cg03183882_red$value)

```

```{r}
library(patchwork)
fig7_c = graph_for_fig
fig7_c_legend = graph1_withLegend



```

```{r}
### Plotting
pdf_export_directory = '~/dnamstability/dnamstability-abutler/pdf_plots/' #where I'm saving figures
pdf(file = paste0(pdf_export_directory, "fig7c_legend.pdf"),   # The directory you want to save the file in
    width = (180/25.4), # The width of the plot in inches (converted from mm)
    height = (110/25.4)) # The height of the plot in inches
fig7_c_legend
dev.off()
###


```




Let's repeat this using SeSAMe and ComBat adjusted FIs...actually let's not. This isn't bias, it's an outlier.
```{r}



```


























Next step: 
Both red/green stdev seem to be an outlier for that one high var subject A sample. Is it predictive of beta value range within-sample? Perhaps we could color clock CpGs by stdev in the above plots.

# Removal of high signal/noise low level data

## Load data and metadata
```{r}
# Load raw data
df_normalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_normalizedBetas_raw_abutler.Rds")

df_unnormalized = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_unnormalizedBetas_raw_abutler.Rds")

SampleAnnotation = readRDS("/home/sagemaker-user/dnamstability/dnamstability-abutler/data/raw_data/20221020_dnamstability_SampleAnnotation_for_ibishof.Rds")

infinium_methylationepic_v_1_0_b5_manifest_file <- read_csv("infinium-methylationepic-v-1-0-b5-manifest-file.csv",
    skip = 7)

df_normalized$X205832330169_R03C01 = NULL
df_unnormalized$X205832330169_R03C01 = NULL

SampleAnnotation$row_number = substr(SampleAnnotation$ChIP_ID,16,16)

SampleAnnotation = SampleAnnotation %>%
  dplyr::select(ChIP_ID,subject,row_number,array,pooled)
SampleAnnotation$XChIP_ID = paste0("X",SampleAnnotation$ChIP_ID)
SampleAnnotation$ChIP_ID=NULL

idat_dir = file.path('/home/sagemaker-user/dnamstability/dnamstability-abutler/data/idat_files/')
rgSet = read.metharray.exp(idat_dir, extended = TRUE)

GrnSD = assay(rgSet, "GreenSD")   # Green SD across beads
RedSD = assay(rgSet, "RedSD")     # Red SD across beads
nBeads = assay(rgSet, "NBeads") 

df_RedSD = as.data.frame(RedSD)
df_GrnSD = as.data.frame(GrnSD)

# although the main point is to flag high SD red and green channel CpGs, I am also going to flag low-n nBeads too
df_nBeads = as.data.frame(nBeads)

# drop outlier from low-level data
# df_RedSD$`205832330169_R03C01` = NULL
# df_GrnSD$`205832330169_R03C01` = NULL
# df_nBeads$`205832330169_R03C01` = NULL

# prune to only relevant addresses (type I and type II probes)
infinium_methylationepic_v_1_0_b5_manifest_file = infinium_methylationepic_v_1_0_b5_manifest_file %>%
  filter(Infinium_Design_Type=="I" | Infinium_Design_Type=="II")

# convert addresses to integer
infinium_methylationepic_v_1_0_b5_manifest_file$AddressA_ID = as.integer(infinium_methylationepic_v_1_0_b5_manifest_file$AddressA_ID
)
infinium_methylationepic_v_1_0_b5_manifest_file$AddressB_ID = as.integer(infinium_methylationepic_v_1_0_b5_manifest_file$AddressB_ID
)
```


I'll use the modified z-score method from Iglewicz and Hoaglin to detect outliers (like the CpG shown above) in either red or green, and remove those from the dataset. I'll need to assess the effect this has on variability. Perhaps visually by comparing heatmaps, and then statistically using ICC or Sample-to-sample distances.

Boris Iglewicz and David Hoaglin (1993), "Volume 16: How to Detect and Handle Outliers", The ASQC Basic References in Quality Control: Statistical Techniques, Edward F. Mykytka, Ph.D., Editor. (as per https://www.itl.nist.gov/div898/handbook/eda/section3/eda35h.htm)

## outlier check
```{r}
detect_outliers <- function(x) {
  # Use the modified Z-score method to detect outliers
  x = as.numeric(x)
  # Calculate median
  med <- median(x)
  # Calculate MAD
  mad <- median(abs(x - med))
  # Calculate modified Z-score
  mzscore <- 0.6745 * (x - med) / mad
  # Return values outside of bounds
  return(abs(mzscore) > 3.5)
}

detect_outliers <- function(x) {
  # Return values outside of bounds
  return(abs(0.6745 * (x - median(x)) / median(abs(x - median(x)))) > 7)
}
```


```{r}
# drop_these will eventually show "TRUE" if any cell has an outlier for the SD data in either channel (typeII) or either address (typeI)
#infinium_methylationepic_v_1_0_b5_manifest_file = infinium_methylationepic_v_1_0_b5_manifest_file[1:100,]



# define the outlier scan function, which will be used to apply detect_outliers() to each CpG
# type_II_manifest_outlier_scan_function <- function(manifest_row) {
#   temp_AddressA_ID <- manifest_row[3]
#   temp_data_M_sd <- df_GrnSD[rownames(df_GrnSD) == temp_AddressA_ID,]
#   temp_data_U_sd <- df_RedSD[rownames(df_RedSD) == temp_AddressA_ID,]
#   test_M = detect_outliers(temp_data_M_sd[1,])
#   test_U = detect_outliers(temp_data_U_sd[1,])
#   return(as.logical(test_M | test_U))
# }

type_II_manifest_outlier_scan_function_green <- function(manifest_row) {
  test_U = detect_outliers(as.numeric(df_GrnSD[rownames(df_GrnSD) == manifest_row[2],][1,]))
  return(as.logical(test_U))
}

type_II_manifest_outlier_scan_function_red <- function(manifest_row) {
  test_U = detect_outliers(as.numeric(df_RedSD[rownames(df_RedSD) == manifest_row[2],][1,]))
  return(as.logical(test_U))
}
```

```{r}
# prune manifest and df_normalized to only type_II probes
type_II_manifest = infinium_methylationepic_v_1_0_b5_manifest_file %>%
  filter(Infinium_Design_Type == "II") %>% select(IlmnID,AddressA_ID)
#troubleshooting
#type_II_manifest = type_II_manifest[1:100,]
```

### (scrapped) based on other samples

#### single core
```{r}


# library(tictoc)
# tic()
# type_II_to_drop_red = apply(type_II_manifest,1,type_II_manifest_outlier_scan_function_red)
# type_II_to_drop_green = apply(type_II_manifest,1, type_II_manifest_outlier_scan_function_green)
# 
# type_II_to_drop_red = data.frame(t(type_II_to_drop_red))
# type_II_to_drop_green = data.frame(t(type_II_to_drop_green))
# type_II_to_drop = (type_II_to_drop_green | type_II_to_drop_red)
# toc()
# sum(type_II_to_drop,na.rm = TRUE)
```

#### parallel note that the above might work better typed in the console rather than run as part of a notebook ###

```{r}
# library(parallel)
# library(tictoc)
# type_II_manifest_t = data.frame(t(type_II_manifest))
# tic()
# # Define the number of cores to use
# num_cores <- detectCores() - 3
# # Apply the function in parallel using mclapply()
# 
# #type_II_to_drop <- mclapply(type_II_manifest_t, type_II_manifest_outlier_scan_function, mc.cores = num_cores-1)
# type_II_to_drop_red <- mclapply(type_II_manifest_t, type_II_manifest_outlier_scan_function_red, mc.cores = num_cores)
# type_II_to_drop_green <- mclapply(type_II_manifest_t, type_II_manifest_outlier_scan_function_green, mc.cores = num_cores)
# 
# 
# 
# # Convert the output to a matrix
# #type_II_to_drop <- do.call(rbind, type_II_to_drop)
# type_II_to_drop_red <- do.call(rbind, type_II_to_drop_red)
# type_II_to_drop_green <- do.call(rbind, type_II_to_drop_green)
# toc()
# type_II_to_drop_red = data.frame(type_II_to_drop_red)
# type_II_to_drop_green = data.frame(type_II_to_drop_green)
# 
# 
# sum(type_II_to_drop_red,na.rm = TRUE)
```

```{r}
# library(foreach)
# library(doParallel)
# type_II_manifest_t = data.frame(t(type_II_manifest))
# num_cores <- detectCores() - 5
# cl <- makeCluster(num_cores)
# 
# type_II_to_drop_red <- foreach(i = 1:length(type_II_manifest_t)) %dopar% {
#   type_II_manifest_outlier_scan_function_red(type_II_manifest_t[[i]])
# }
# type_II_to_drop_green <- foreach(i = 1:length(type_II_manifest_t)) %dopar% {
#   type_II_manifest_outlier_scan_function_green(type_II_manifest_t[[i]])
# }
# 
# stopCluster(cl)
```

### based on low-level signal to noise
```{r}
grn_matrix = as.matrix(df_GrnSD)/as.matrix(df_Green)
red_matrix = as.matrix(df_RedSD)/as.matrix(df_Red)

df_grnSignalToNoise = data.frame(grn_matrix)
df = reshape2::melt(df_grnSignalToNoise)
df$green = reshape2::melt(df_Green)[,2]
df$green_SD = reshape2::melt(df_GrnSD)[,2]
colnames(df)[2] = "FI_stdev/FI"

# histogram showing distribution of FI/FI_Stdev for green channel
graph <- ggplot(df, aes(x = `FI_stdev/FI`)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.01) +
  theme_bw() + 
  ggtitle("Distribution of FI_Stdev/FI for green channel") 
graph

source("group_means_shiny.r")
df = na.omit(df)
#group_means_shiny(df)


df_redSignalToNoise = data.frame(red_matrix)
df = reshape2::melt(df_redSignalToNoise)
df$red = reshape2::melt(df_Red)[,2]
df$red_SD = reshape2::melt(df_RedSD)[,2]
colnames(df)[2] = "FI_stdev/FI"

# histogram showing distribution of FI/FI_Stdev for red channel
graph <- ggplot(df, aes(x = `FI_stdev/FI`)) +
  geom_histogram(position = 'identity', alpha = 0.8, binwidth = 0.01) +
  theme_bw() + 
  ggtitle("Distribution of FI_Stdev/FI for red channel")
graph
#ggplotgui::ggplot_shiny(df)

source("group_means_shiny.r")
df = na.omit(df)
#group_means_shiny(df)

df_grnSignalToNoise$address = rownames(df_grnSignalToNoise)
df_redSignalToNoise$address = rownames(df_redSignalToNoise)

# make a unique_ID to join columns from red and green df
df_meltRed = reshape2::melt(df_redSignalToNoise,id.vars="address")
df_meltRed$unique_ID = paste0(df_meltRed$address,df_meltRed$variable)

df_meltgrn = reshape2::melt(df_grnSignalToNoise,id.vars="address")
df_meltgrn$unique_ID = paste0(df_meltgrn$address,df_meltgrn$variable) # these are identical. Just going to join dfs

df_melt = data.frame(address = df_meltRed$address,
                     sample = df_meltRed$variable,
                     `red_FI_stdev/FI` = df_meltRed$value,
                     red_FI = reshape2::melt(df_Red)[,2],
                     red_SD = reshape2::melt(df_RedSD)[,2],
                     `green_FI_stdev/FI` = df_meltgrn$value,
                     green_FI = reshape2::melt(df_Green)[,2],
                     green_SD = reshape2::melt(df_GrnSD)[,2]
                     )

graph <- ggplot(df_melt, aes(x = green_FI, y = red_FI_stdev.FI)) +
  geom_point() +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
#graph

df_melt_na = na.omit(df_melt)

#group_means_shiny(df_melt_na)

sum(df_melt_na$red_FI_stdev.FI==df_melt_na$green_FI_stdev.FI)
```



Set a cutoff
```{r}
# cutoff = X
cutoff = 0.5 # any probe who's CV is >0.3 gets converted to NA below

grn_matrix = data.frame((as.matrix(df_GrnSD)/as.matrix(df_Green)) > cutoff)
grn_matrix$rownames = rownames(grn_matrix)

red_matrix = data.frame((as.matrix(df_RedSD)/as.matrix(df_Red)) > cutoff)
red_matrix$rownames = rownames(red_matrix)

#type_II_names = infinium_methylationepic_v_1_0_b5_manifest_file$Name[infinium_methylationepic_v_1_0_b5_manifest_file$Infinium_Design_Type=="II"]

type_II_names = infinium_methylationepic_v_1_0_b5_manifest_file %>%
  filter(Infinium_Design_Type=="II")

type_I_red_names = infinium_methylationepic_v_1_0_b5_manifest_file %>%
  filter(Infinium_Design_Type=="I" & infinium_methylationepic_v_1_0_b5_manifest_file$Color_Channel=="Red")

type_I_green_names = infinium_methylationepic_v_1_0_b5_manifest_file %>%
  filter(Infinium_Design_Type=="I" & infinium_methylationepic_v_1_0_b5_manifest_file$Color_Channel=="Grn")

# lists of cpgs
red_matrix_type_II = red_matrix %>%
  filter(rownames %in% type_II_names$AddressA_ID)
grn_matrix_type_II = grn_matrix %>%
  filter(rownames %in% type_II_names$AddressA_ID)

red_matrix_type_I = red_matrix %>%
  filter(rownames %in% type_I_red_names$AddressA_ID | rownames %in% type_I_red_names$AddressB_ID)

grn_matrix_type_I = grn_matrix %>%
  filter(rownames %in% type_I_green_names$AddressA_ID | rownames %in% type_I_green_names$AddressB_ID)

```

```{r}
to_drop = c()
for (i in 1:length(type_II_names$Name)){
  if (x == 2){}
}
```



```{r}
# prune manifest and df_normalized to only type_I probes
type_I_red_manifest = infinium_methylationepic_v_1_0_b5_manifest_file %>%
  filter(Infinium_Design_Type == "I" & Color_Channel=="Red")

type_I_red_manifest_t = data.frame(t(type_I_red_manifest))

type_I_red_manifest_outlier_scan_function <- function(manifest_row) {
  temp_AddressA_ID <- manifest_row[3]
  temp_AddressB_ID <- manifest_row[5]
  temp_data_M_sd = df_RedSD[rownames(df_RedSD)==temp_AddressB_ID,]
  temp_data_U_sd = df_RedSD[rownames(df_RedSD)==temp_AddressA_ID,]
  test_M = detect_outliers(temp_data_M_sd[1,])
  test_U = detect_outliers(temp_data_U_sd[1,])
  return(as.logical(test_M | test_U))
}

type_I_red_to_drop <- mclapply(type_I_red_manifest_t, type_I_red_manifest_outlier_scan_function, mc.cores = num_cores)
type_I_red_to_drop <- do.call(rbind, type_I_red_to_drop)
type_I_red_to_drop = data.frame(type_I_red_to_drop)
```

```{r}
# prune manifest and df_normalized to only type_I probes
type_I_green_manifest = infinium_methylationepic_v_1_0_b5_manifest_file %>%
  filter(Infinium_Design_Type == "I" & Color_Channel=="Grn")

type_I_green_manifest_t = data.frame(t(type_I_green_manifest))

type_I_green_manifest_outlier_scan_function <- function(manifest_row) {
  temp_AddressA_ID <- manifest_row[3]
  temp_AddressB_ID <- manifest_row[5]
  temp_data_M_sd = df_GrnSD[rownames(df_GrnSD)==temp_AddressB_ID,]
  temp_data_U_sd = df_GrnSD[rownames(df_GrnSD)==temp_AddressA_ID,]
  test_M = detect_outliers(temp_data_M_sd[1,])
  test_U = detect_outliers(temp_data_U_sd[1,])
  return(as.logical(test_M | test_U))
}

type_I_green_to_drop <- mclapply(type_I_green_manifest_t, type_I_green_manifest_outlier_scan_function, mc.cores = num_cores-1)
type_I_green_to_drop <- do.call(rbind, type_I_green_to_drop)
type_I_green_to_drop = data.frame(type_I_green_to_drop)
```

```{r}
to_drop = rbind(type_II_to_drop, type_I_red_to_drop, type_I_green_to_drop)
rownames(to_drop) = c(type_II_manifest$IlmnID,type_I_red_manifest$IlmnID,type_I_green_manifest$IlmnID)
```




## scrap code below

```{r}

type_II_to_scan = df_normalized %>%
  filter(rownames(df_normalized) %in% type_II_manifest$IlmnID)

drop_these_type_II = drop_these %>%
  filter(rownames(drop_these) %in% type_II_manifest$IlmnID)





# drop_these should have the same dimensions as df_normalized here, and we'll truncate it down to specific subsets before rejoining
drop_these = matrix(nrow=nrow(df_normalized),
                    ncol = ncol(df_normalized))
drop_these = data.frame(drop_these)
colnames(drop_these) = colnames(df_normalized)
rownames(drop_these) = rownames(df_normalized)
#initialize df, will eventually be logical





# sort to match the rows
sort_to_match_rows = function(df_to_sort,template_df){
  idx = match(rownames(df_to_sort),rownames(template_df)) # rows
  df_to_sort = df_to_sort[order(idx),]
  return(df_to_sort)
}

type_II_manifest = as.data.frame(type_II_manifest) # a tibble, convert to df

rownames(type_II_manifest) = type_II_manifest$IlmnID # give it row names corresponding to IlmnID (cgXXX)


type_II_manifest = sort_to_match_rows(type_II_manifest,drop_these_type_II) # sort
sum(rownames(type_II_manifest)==rownames(drop_these_type_II)) # Check rownames the same:
sum(rownames(type_II_manifest)==rownames(type_II_to_scan)) # Check rownames the same:

number_of_observations = length(type_II_to_scan) # count number of observations
type_II_to_scan$AddressA_ID = type_II_manifest$AddressA_ID # add the address

type_II_outlier_scan_function <- function(vector,rowname) {
  temp_AddressA_ID <- vector[number_of_observations + 1]
  temp_data_M_sd <- df_GrnSD[rownames(df_GrnSD) == temp_AddressA_ID,]
  temp_data_U_sd <- df_RedSD[rownames(df_RedSD) == temp_AddressA_ID,]
  test_M = detect_outliers(temp_data_M_sd[1,])
  test_U = detect_outliers(temp_data_U_sd[1,])
  return(as.logical(test_M | test_U))
}


drop_these_type_II = data.frame(drop_these_type_II)
for (i in 1:length(type_II_to_scan$X205832330169_R01C01)){
  drop_these_type_II[i,] = type_II_outlier_scan_function(type_II_to_scan[i,],rownames(type_II_to_scan[i,]))
}

drop_these_type_II <- apply(type_II_to_scan, 1, function(x) type_II_outlier_scan_function(x, rownames(x)))



```

```{r}
for(i in 1:length(rownames(drop_these))){
  
  if (temp_meta$Infinium_Design_Type[1] == "II"){
    temp_data_M_sd = df_GrnSD[rownames(df_GrnSD)==temp_meta$AddressA_ID,]
    temp_data_U_sd = df_RedSD[rownames(df_GrnSD)==temp_meta$AddressA_ID,]
    test_M = detect_outliers(temp_data_M_sd[1,])
    test_U = detect_outliers(temp_data_U_sd[1,])
    drop_these[i,] = test_M | test_U
  }
}


for(i in 1:length(rownames(drop_these))){
  temp_meta = infinium_methylationepic_v_1_0_b5_manifest_file[infinium_methylationepic_v_1_0_b5_manifest_file$IlmnID == cpgs[i],]
  if (temp_meta$Infinium_Design_Type[1] == "II"){
    temp_data_M_sd = df_GrnSD[rownames(df_GrnSD)==temp_meta$AddressA_ID,]
    temp_data_U_sd = df_RedSD[rownames(df_GrnSD)==temp_meta$AddressA_ID,]
    test_M = detect_outliers(temp_data_M_sd[1,])
    test_U = detect_outliers(temp_data_U_sd[1,])
    drop_these[i,] = test_M | test_U
  }
}

for(i in 1:length(rownames(drop_these))){
  #Type I probes have two probes or addresses (A and B) corresponding to unmethylated and methylated probes, respectively (UMtools documentation)
  if (temp_meta$Infinium_Design_Type[1] == "I" & temp_meta$Color_Channel=="Red"){
    temp_data_M_sd = df_RedSD[rownames(df_RedSD)==temp_meta$AddressB_ID,]
    temp_data_U_sd = df_RedSD[rownames(df_RedSD)==temp_meta$AddressA_ID,]
    test_M = detect_outliers(temp_data_M_sd[1,])
    test_U = detect_outliers(temp_data_U_sd[1,])
    drop_these[i,] = test_M | test_U
  }
}

for(i in 1:length(rownames(drop_these))){
    if (temp_meta$Infinium_Design_Type[1] == "I" & temp_meta$Color_Channel=="Grn"){
    temp_data_M_sd = df_GrnSD[rownames(df_GrnSD)==temp_meta$AddressB_ID,]
    temp_data_U_sd = df_GrnSD[rownames(df_GrnSD)==temp_meta$AddressA_ID,]
    test_M = detect_outliers(temp_data_M_sd[1,])
    test_U = detect_outliers(temp_data_U_sd[1,])
    drop_these[i,] = test_M | test_U
  }
}


```
